Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-09 paper.tex
Took a stab at some related work and a conclusion.

\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S', (\mastate',\mstore') \notin S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

@@

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control, given in figure \ref{fig:shift-reset} (adapted from
~\citep{ianjohnson:Biernacki2006274}). We could view each function call
as inserting a prompt, and returns as aborting to the nearest prompt.

@@

Turning this into a table-based semantics involves making prompts a
point of indirection, just like function calls. Memoization also gets
a new context to consider, calling a continuation, because composable
continuations act like functions. Figure \ref{fig:shift-reset-table0}
has what one might naturally write using just this
information. Unfortunately, since $\rt$ frames contain a store, and
continuations can now appear in the store, this introduces a
circularity that could cause the analysis to never terminate. For
example, a loop that continually captures continuations would
introduce unboundedly many continuation values due to the ever-growing
store. This means the store is not finite height, and there may be no
fixed point.

\paragraph{Possible fixes to non-termination:} The simplest fix to the
circularity would be to strip stores out of the context of the $\rt$
frame in a captured continuation, and additionally consider contexts
with a stripped store an abstraction of the same context with any
store. This is a fairly brutal approximation of a captured
continuation, so an alternative is to make this tunable with our
precision-tuning friend, $\alloc$. Then, instead of entirely removing
the store component of the $\rt$ context, we can replace it with an
address of possible stores that it would approximate. The first
brute-force approach is an instantiation of this, where the allocated
address is always the same.

@@

The new indirection possibility changes $\Context$ to also include
$\maddr$ where there was previously a $\mstore$, and $\KTab$ now
additionally maps $\Addr$ to a set of stores. We introduce three new
metafunctions to facilitate the new approximation:

\newcommand{\replacectx}{\mathit{replace}\mstore}
\newcommand{\addstore}{\mathit{add}\mstore}
\begin{align*}
  \returns(\mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mktab(\maddr)}
  \\[2pt]
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \mstore), \mval) &=
    \joinone{\mmemo}{((\mexpr,\menv), \mstore)}{\mval} \\
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \maddr), \mval) &=
    \mmemo \sqcup \bigsqcup\limits_{\mstore \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore)}{\mval}
  \\[2pt]
  \approximate(\mkont, \mktab, \maddr) &= (\replacectx(\mkont), \addstore{\mkont}) \\
  \text{where }
   \replacectx(\krt{((\mexpr,\menv),\_)}) &= \krt{((\mexpr,\menv),\maddr)} \\
   \replacectx(\kar{\mexpr,\menv,\mkont}) &= \kar{\mexpr,\menv,\replacectx(\mkont)} \\
   \replacectx(\kfn{\mval, \mkont}) &= \kfn{\mval,\replacectx(\mkont)}
  \\[2pt]
   \addstore(\krt{((\mexpr,\menv),\mstore)}) &= \joinone{\mktab}{\maddr}{\mstore} \\
   \addstore(\krt{((\mexpr,\menv),\maddr')}) &= \joinm{\mktab}{\maddr}{\mktab(\maddr')} \\
   \addstore(\kar{\mexpr,\menv,\mkont}) &= \addstore(\mkont) \\
   \addstore(\kfn{\mval, \mkont}) &= \addstore(\mkont)
\end{align*}

@@

% Since continuations can now appear in the store, and the store is in
% the $\rt$ continuation, we introduced a circularity in the data
% represented in our machine, which can lead to a non-terminating
% analysis --- this is precisely what the AAM approach avoids. There is
% currently no known model of computation that allows stack capture and
% replacement/composition that has decidable reachability, so we choose
% to approximate in this case. We represent only a bounded amount of a
% continuation, possibly with some gaps, after which we say that any
% continuation matching that ``outline'' is acceptable. The amount to
% represent is a tunable parameter; we can represent the continuation up
% to boundedly-many function calls, or have innermost prompt tags
% matching $\many\mtag$ for boundedly-long $\many\mtag$ (or a mix of
% $dw$ frames and prompt tags), etc.

@@

AAM, PDCFA, CFA2, CFA2+, Reps

@@

Tables, truncation and prompts are the solution to all life's ills.

>>>>

\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

@@

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control, given in figure \ref{fig:shift-reset} (adapted from
~\citep{ianjohnson:Biernacki2006274}). We could view each function
call as inserting a prompt, and returns as aborting to the nearest
prompt. The changed semantic spaces for the shift/reset semantics are as
follows:

\begin{align*}
  \mstate \in \State &::= \tpl{\mpoint, \mstore, \mkont, \mmkont} \\
  \mpoint \in \Point &::= (\mexpr, \menv) \mid \mval \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \mid \vcomp{\mkont} \\
  \mmkont \in \MKont &::= \kmt \mid \mkapp{\mkont}{\mmkont}
\end{align*}

@@

Turning this into a table-based semantics involves making prompts a
point of indirection, just like function calls. Memoization also gets
a new context to consider, calling a continuation, because composable
continuations act like functions. The new semantic spaces are then

\begin{align*}
  \mctx \in \Context &::= ((\mexpr,\menv),\mstore) \mid (\vcomp{\mkont},\mval,\mstore) \\
  \mkont \in \Kont &= \kmt \mid \krt{\mctx} \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \mmkont \in \MKont &::= \kmt \mid \kprompt{\mctx} \\
  \mktab \in \KTab &= \Context \to (\wp(\Kont \times \MKont) \cup \Kont) \\
  \mmemo \in \Memo &= \Context \to \wp(\Value)
\end{align*}

Figure \ref{fig:shift-reset-table0} has what one might naturally write
using just this information. Unfortunately, since $\rt$ frames contain
a store, and continuations can now appear in the store, this
introduces a circularity that could cause the analysis to never
terminate. For example, a loop that continually captures continuations
would introduce unboundedly many continuation values due to the
ever-growing store. This means the store is not finite height, and
there may be no fixed point.

\paragraph{Possible fixes to non-termination:} The simplest fix to the
circularity would be to strip stores out of the context of the $\rt$
frame in a captured continuation, and additionally consider contexts
with a stripped store an abstraction of the same context with any
store. This is a fairly brutal approximation of a captured
continuation, so an alternative is to make this tunable with our
precision-tuning friend, $\alloc$. Then, instead of entirely removing
the store component of the $\rt$ context, we can replace it with an
address of possible stores that it would approximate.

@@

The new indirection possibility changes $\Context$ to also include
$\maddr$ where there was previously a $\mstore$ (though contexts for
continuation calls remain unchanged), and $\KTab$ now additionally
maps $\Addr$ to a set of stores. The rules that change are presented in figure \ref{fig:shift-reset-table1}.

\begin{align*}
  \mctx \in \Context &::= ((\mexpr,\menv),\mstore) \mid ((\mexpr,\menv),\maddr) \mid (\vcomp{\mkont},\mval,\mstore) \\
  \msctx \in \SContext &::= ((\mexpr,\menv),\maddr) \\
  \mskont \in \SKont &= \krt{\msctx} \mid \ldots \\
  \mval \in \Value &= \vclo{\slam{\mvar}{\mexpr}}{\menv} \mid \vcomp{\mskont} \\
  \mmkont \in \MKont &::= \kmt \mid \kprompt{\mctx} \\
  \mktab \in \KTab &= (\Context \to (\wp(\Kont \times \MKont) \cup \Kont)) \\
                   &\cup (\Addr \to \wp(\Store))
\end{align*}

@@

The metafunctions mentioned in the fixed semantics all deal with the
addition of $\maddr$ to contexts. If the context in an $\rt$ frame is
approximate, we must return to all the continuations known for all the
contexts it approximates ($\returns$). At return boundaries, the memo
table must add the result to all the represented contexts
($\memoize$). At capture time, we strip the store in the $\rt$ frame
if there is one, and replace it with an address.

\newcommand{\replacectx}{\mathit{replace}\mstore}
\newcommand{\addstore}{\mathit{add}\mstore}
\begin{align*}
  \returns(\mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mktab(\maddr)}
  \\[2pt]
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \mstore), \mval) &=
    \joinone{\mmemo}{((\mexpr,\menv), \mstore)}{\mval} \\
  \memoize(\mmemo, \mktab, ((\mexpr,\menv), \maddr), \mval) &=
    \mmemo \sqcup \bigsqcup\limits_{\mstore \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore)}{\mval}
  \\[2pt]
  \approximate(\mkont, \mktab, \maddr) &= (\replacectx(\mkont), \addstore{\mkont}) \\
  \text{where }
   \replacectx(\kmt) &= \kmt \\
   \replacectx(\krt{((\mexpr,\menv),\_)}) &= \krt{((\mexpr,\menv),\maddr)} \\
   \replacectx(\kar{\mexpr,\menv,\mkont}) &= \kar{\mexpr,\menv,\replacectx(\mkont)} \\
   \replacectx(\kfn{\mval, \mkont}) &= \kfn{\mval,\replacectx(\mkont)}
  \\[2pt]
   \addstore(\kmt) &= \mktab \\
   \addstore(\krt{((\mexpr,\menv),\mstore)}) &= \joinone{\mktab}{\maddr}{\mstore} \\
   \addstore(\krt{((\mexpr,\menv),\maddr')}) &= \joinm{\mktab}{\maddr}{\mktab(\maddr')} \\
   \addstore(\kar{\mexpr,\menv,\mkont}) &= \addstore(\mkont) \\
   \addstore(\kfn{\mval, \mkont}) &= \addstore(\mkont)
\end{align*}

The result of this viewpoint is a sound, precise, and computable
semantics for a ``pushdown approach'' to analyzing delimited,
composable control. To handle non-composable control, we can simply
remove the prompts and keep $\approximate$.

@@

The immediately related work is that of PDCFA
\citet{ianjohnson:earl2010pdcfa},
CFA2~\citet{ianjohnson:vardoulakis-lmcs11,
  ianjohnson:Vardoulakis2011Pushdown}, and AAM
~\citet{dvanhorn:VanHorn2010Abstracting}, the first two of which we
recreated in full detail. The version of CFA2 that handles
\rackett{call/cc} does not handle composable control, and is dependent
on a restricted CPS representation. They also had no way to tune the
precision of their first class continuation approximation. Since
\rackett{call/cc} can be simulated with shift/reset, this work
supercedes theirs.

Pushdown models have existed in the first-order static analysis
literature ~\citet[Chapter
7]{local:muchnick:jones:flow-analysis:1981}\citet{ianjohnson:reps:pushdown:1995},
and the first-order model checking literature
\citet{ianjohnson:bouajiani:esparza:pushdown:1997}, for some time. The
higher-order setting imposes additional challenges that make their
methods difficult to adapt. The most important constraint is that we
can't know all call-sites of a function/method before the analysis
begins, which their methods heavily rely on.

@@

As the programming world continues to embrace behavioral values, it
becomes more important to import the powerful techniques pioneered by
the first-order analysis and model checking communities. CFA2 and
PDCFA paved the way, and in large part inspired this work. It is our
view that systematic approaches to applying the techniques are pivotal
to scaling them to ``real languages.'' We believe that the recipe that
this paper set forth is a step in that direction. That is, make
continuation tables keyed with enough context, and memoize at the
introduced indirection points. The result in a language with
well-bracketed control is a ``pushdown analysis'' using
summarization. In a language without well-bracketed control, we are
not chained to a pushdown automaton as the target of the
approximation, so the techniques still apply and give better precision
than regular methods.
